---
title: "SynPUF-propensity-model"
author: "Erika Braithwaite"
date: '2018-05-23'
output:
  html_document: default
  github_document: default
---

```{r setup, echo = FALSE}
pacman::p_load(aws.s3, magrittr, dplyr, MatchIt, ggplot2, plotly, tableone, forcats, tidyr, knitr, kableExtra, gridExtra, tibble, stringr, markdown)

dat = readRDS('study_population_6m.RData')

logistic.regression.or.ci <-function(regress.out, level = 0.95)
{ pacman::p_load(gtools)
        usual.output<-summary(regress.out)
        regress.ci <- confint.default(regress.out)
        
        dfBci <- data.frame(usual.output$coefficients[,1], usual.output$coefficients[,2], regress.ci,
                            usual.output$coefficients[,3], usual.output$coefficients[,4])
        colnames(dfBci) <- c("Estimate", "SE", "Lower CI", "Upper CI", "z value", "Pr(>|z|)")
        dfBci[] <-lapply(dfBci,formatC,digits=2,format="f")
        
        #get OR ci & p-value 
        slopes.ci <- regress.ci[-1,]
        OR.ci<-exp(slopes.ci)
        p.value =  usual.output$coefficients[,4]
        star = stars.pval(p.value)
        
        #make table of OR and OR ci & add p-value 
        e.coeff <-exp(usual.output$coefficients[,1])
        e.ci <-exp(regress.ci)
        
        dfORci <-data.frame(e.coeff, e.ci, p.value) 
        colnames(dfORci) <- c("OR", "Lower CI", "Upper CI", 'p-value')
        dfORci[] <-lapply(dfORci,formatC,digits=2)
        dfORci2 <- cbind(dfORci, star)
        dfORci2 <- dfORci2[-1,]
        output <- list(regression.table.ci = dfBci, OR.table = dfORci2)
        
        return(output)
}
```


First select individuals above 40 at the time they first showed up in the dataset; then exclude anyone died within one year; then select individuals with statin prescription after 2009-01-01 (set first prescription date as cohort entry date); then exclude anyone who took lipid-lowering drugs before 2009-01-01; also exclude anyone who had dialysis one year before cohort entry

```{r data-cleanup, echo = FALSE}
dat$statin = dat$initial_level
dat$statin[dat$statin == 'low'] = '0'
dat$statin[dat$statin == 'high'] = '1'
dat$statin= as.numeric(dat$statin)


catVars = c('hypertension', 'hypercholesterol', 'heart.failure', 
           'PVD', 'injury.poison', 'hospital', 'lab.test')

dat %<>%
  mutate_at(catVars, funs(factor(.)), labels = c('No', 'Yes'))

dat$acute.kidney = dat$outcome
dat$acute.kidney = factor(dat$acute.kidney, labels = c('No readmission', 'Readm. kidney'))

dat$sex = dat$BENE_SEX_IDENT_CD
dat$sex = factor(dat$sex, labels = c('Male', 'Female'))

dat$race = dat$BENE_RACE_CD
dat$race = factor(dat$race, labels = c('White', 'Black', 'Others', 'Hispanic'))



```

```{r reduce-data, echo = FALSE}
dat  %<>% select(DESYNPUF_ID, sex, race, age_entry, hypertension, hypercholesterol, 
                 heart.failure, PVD, injury.poison, hospital, lab.test, statin, acute.kidney) %>% 
  na.omit
```


### Table 1. Covariate balance in total sample N = 839

```{r Create unmatched table 1, echo = FALSE, results = 'hide'}
unmatched_table1 = CreateCatTable(strata = 'statin', data = dat, vars = catVars, test = FALSE, smd = T)
un_table1 = print(unmatched_table1, smd = TRUE)

```

```{r Print unmatched table, echo = FALSE}
kable(un_table1, align = 'c', digits = 2, col.names = c('low statin', 'high statin', 'SMD'))
```

### Estimate propensity score with nearest neighbor 1:1 matching 


```{r Propensity score}
model = matchit(statin ~ hypertension + hypercholesterol + heart.failure + 
                injury.poison + hospital + lab.test, method = 'nearest', data = dat)

kable(model$nn)

```


### Pruned data

```{r Create-matched-table1, echo = F, results = 'hide'}
# Make data frame containing only matched observations
matched_data = match.data(model)

# Matched table of descriptives 
matched_table1 = CreateTableOne(strata = 'statin', data = matched_data, vars = catVars, test = FALSE)

# Include the table with the standardized mean difference
m_table1 = print(matched_table1, smd = TRUE)

```

```{r Matched-table1, echo = FALSE}
kable(m_table1, digits = 2, align = 'c', col.names = c('low statin', 'high statin', 'SMD'))
```

### Comparing SMD between matched and unmatched sample 

```{r Create-comparison-table, echo = F}
big.table = data.frame(variable = matched_table1$MetaData$vars, 
                                        Unmatched = ExtractSmd(unmatched_table1), 
                                        Matched = ExtractSmd(matched_table1))



names(big.table) = c('variable', 'unmatched sample', 'matched sample')
rownames(big.table) <- c()
kable(big.table, digits = 2, align = 'c')
```

### Distribution of standardized mean differences before and after matching

```{r distribution-plot-matched, echo = FALSE}
plot.data = big.table %>% 
  gather(key = sample, value = smd, `unmatched sample`:`matched sample`) %>% 
  mutate(sample = factor(sample), 
         smd = round(smd, 2))


ggplot(data = plot.data, aes(x = variable, y = smd, group = sample, color = sample)) +
        geom_point() +
        geom_hline(yintercept = 0.05, color = "black", size = 0.1) +
        coord_flip() +
        theme_bw() + theme(legend.key = element_blank())
```

### Distribution of propensity scores in matched and unmatched sample

```{r distribution-plot-unmatched, echo = FALSE}

prop.total = glm(statin ~ hypertension + hypercholesterol + heart.failure + 
                injury.poison + hospital + lab.test, family = binomial, data = dat)

prop.df = data.frame(distance = predict(prop.total, type = 'response'), 
                     statin = prop.total$model$statin)


pscore.plot.unmatched = ggplot(data = prop.df, aes(x = distance, fill = as.factor(statin))) + 
  geom_density(alpha = 0.5) + 
  ggtitle('Distribution of propensity scores in unmatched sample') + 
  labs(x = 'propensity score', 
       y = 'frequency') + 
  scale_fill_discrete(labels = c('high statin', 'low statin'), 
                      guide_legend(title = 'Treatment'))

pscore.plot.matched = ggplot(data = matched_data, aes(x = distance, fill = as.factor(statin))) + 
        geom_density(alpha = 0.5) + 
        ggtitle('Distribution of propensity scores in matched sample') + 
  labs(y = 'frequency', 
       x = 'propensity score') + 
  scale_fill_discrete(labels = c('high statin', 'low statin'), 
                      guide_legend(title = 'Treatment'))

grid.arrange(pscore.plot.unmatched, pscore.plot.matched)
```

### Outcome models

```{r Create outcome models, echo = FALSE}
crude = glm(acute.kidney ~ statin, data = dat, family = binomial(link = 'logit')) %>% 
         logistic.regression.or.ci() %>% 
         extract2(2) %>% 
         rownames_to_column() %>% 
         filter(str_detect(rowname, 'statin')) %>% # select covariates
         mutate(Variable = 'statin', 
                Model = "Crude") %>% # Model name
         mutate_at(c('OR', 'Lower CI', 'Upper CI'),funs(as.numeric))
 
 adjusted = glm(acute.kidney ~ statin + hypertension + hypercholesterol + heart.failure + 
                  injury.poison + hospital + lab.test, data = dat, family = binomial(link = 'logit')) %>% 
         logistic.regression.or.ci() %>% 
         extract2(2) %>% 
         rownames_to_column() %>% 
         filter(str_detect(rowname, 'statin')) %>% # select covariates
         mutate(Variable = 'statin', 
                Model = "Adjusted") %>% # Model name
         mutate_at(c('OR', 'Lower CI', 'Upper CI'),funs(as.numeric))
 
 matched = glm(acute.kidney ~ statin + hypertension + hypercholesterol + heart.failure + 
                 injury.poison + hospital + lab.test, data = matched_data, 
               family = binomial(link = 'logit')) %>% 
         logistic.regression.or.ci() %>% 
         extract2(2) %>% 
         rownames_to_column() %>% 
         filter(str_detect(rowname, 'statin')) %>% # select covariates
         mutate(Variable = 'statin', 
                Model = "Matched") %>% # Model name
         mutate_at(c('OR', 'Lower CI', 'Upper CI'),funs(as.numeric))
 
 model.compare = data.frame(rbind(crude, adjusted, matched)) %>% 
         arrange(Variable)
 
 # Reorder the Model levels so they appear in the legend in the order I want
model.compare$Model = factor(model.compare$Model,levels(factor(model.compare$Model))[c(2,1,3)])
```


```{r Plot-models, echo = FALSE}

model.compare$X = c(3,2,1)

ggplot(data = model.compare, aes(y = OR, colour = Model )) + 
        geom_hline(yintercept = 1, colour = grey(1/2)) + 
        coord_flip() + 
        geom_pointrange(aes(x = X, y = OR, ymin = Lower.CI, 
                            ymax = Upper.CI), 
                        lwd = 1/2, position = position_dodge(width = 1/2), 
                        shape = 21, fill = 'WHITE') + 
        ggtitle('Association between high vs low statin use and acute kidney disease readmission') + 
        labs(y = 'Odds Ratio (95% CI)', 
             x = ' ') + 
        scale_y_continuous(trans = "log", limits = c(0.3, 2), 
                           breaks = c(0.3, 0.5, 0.7, 0, .5, 1, 1.5, 2)) + 
  theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.title.y=element_blank()) 
```

```{r knittohtml, echo = FALSE}
markdownToHTML('SynPUF-propensity-model.md', 'SynPUF-propensity-model.html', options=c("use_xhml"))
```

